---
output:
  html_document: default
  pdf_document: default
  word_document: default
---
Analysis of Elections in India
========================================================
By Mohd Hassaan
-------------

## Introduction

India is the largest democracy in the world, it has mainly two kind of election state and national election. National election is conducted on 545 seats. more than 500 million people voted in the 2014 election. What could be more intresting than analysing the data of indian elections through out the years.

In this project we are going to analyse the data of India's National elections from 1977 to 2014.

> **Source**:  https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/26526


### Initial Exploration 
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Loading all the required packages
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(gridExtra)
library(extrafont)
library(RColorBrewer)

#setting the working directory
setwd("C:\\Users\\mohdha\\Documents\\GitHub\\India-Election-Analysis-with-R")

```

### Structure of the data

```{r echo=FALSE, Load_the_Data}

# Load the Data
national_election = read.table('nationalelection.tab',sep="\t", header=TRUE)

# looking the structure of data
str(national_election)

```

In the election data set we have state name(st_name), constituency name(pc_name), total electorates(electors), votes polled(totvotpoll), cadidate name(cand_name) etc.

### Adding new variables

Before we go into any analysis we are going to add some new variables in the data. We are going to add a variable name proportion which gives us the proportion of vote a candidate get. the another variable is booleean variable which store the information about the candidate's winning or loosing.


```{r echo=FALSE, making_variables}

#Making a variable proportion to reflect the vote proportion of each candidate
national_election$proportion = national_election$totvotpoll / national_election$electors

head(national_election)

#making a variable with the winning vote
national_election.winner_vote <- national_election %>%
  group_by(st_name, year,pc_no) %>%
  summarise(win_vote = max(totvotpoll))

#making the left join to the orignal data
Nat_election <- 
  merge(national_election,national_election.winner_vote, by = c('st_name','year','pc_no'),x.all = TRUE )


#making our won variable to reflect the winning status of the candidate
Nat_election$won = ifelse(Nat_election$totvotpoll == Nat_election$win_vote,TRUE, FALSE)

summary(Nat_election)


```

### Cleaning the data

Above we can see that in data we can see that the minimum winning vote is 0. which cant be true.These may not be the wrong values, because manier times candidate was cancelled after suspection of booth capturing, manier times result Whithheal By Courts etc. These observation are very few(5 to be exact) ,but this can effect out analysis so we have to remove these from our data set as a outliers.

```{r echo=FALSE, remove_outliars}
#filtering the outliars
Nat_election <- subset(Nat_election, win_vote != 0)
sprintf('Nation election data has %s Observation after filtering out Outliars.',NROW(Nat_election))

```

```{r echo=FALSE, updated_summary}

#updated summary
summary(Nat_election$win_vote)

```
# Univariate Plots Section

Now We are going to analyse out data in one dimention.

## 1. No of contestant contesting every election year
```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots}

#converting year variable into type factor
Nat_election$year <- as.factor(Nat_election$year)

#plotting the no of candidates contested each year in election
ggplot(Nat_election, aes(x=year)) + 
  geom_histogram(stat = 'count',fill = '#C51162',color="black") +
  theme(axis.text.x=element_text(angle=90, size=9)) +
  coord_flip() + theme_minimal() + scale_y_continuous(breaks=seq(0, 15000, 1000))+
  ggtitle("No of Contestant each year (National)")


```

Here we can see, there is a consistent increase in number of contestant participating in the elction and goes up to 14000 in 1996.But after that it comes to less than 5000 in 1998 and it decreased even further in 1999.

The increase in 1996 could be because so many independent cadidate contested as it was the first election after india became an open economy which give indian citzen too many opportunities and contesting election was one of that.

## 2.Number of contestant based on gender.  

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots_sex}


#plotting the no of candidates contested by gender
ggplot(subset(Nat_election,cand_sex != 'NULL'), aes(x=cand_sex)) + 
  geom_bar(stat = 'count',fill = '#C51162',color="black") +
  geom_text(stat="count", color="black", vjust = -0.1, size=3, aes(y=..count.., label=..count..)) + 
  theme(axis.text.x=element_text(angle=90, size=9)) +
   theme_minimal() +
  ggtitle("Contestant by Gender (National)")


```

We can see how men easily out numbered the women in politics.womens had contested almost less than 5 percent as compared to mens. The other sex(or LGBT) has a long way to go in the indian politics as far as the numbers are concern.
It will be more intresting to investigate how many women out of the few who contested did well in the elections. 

## 3.Number of contestant based on State.

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots_state}

#plots the histogram to see the no of contestant from each state
ggplot(Nat_election, aes(x=st_name)) + 
  geom_histogram(stat = 'count',fill = '#C51162',color="black") +
  theme(axis.text.x=element_text(angle=90, size=9)) +
  coord_flip() + theme_minimal() +
  ggtitle("No of Contestant from each state (National)") + xlab("States") + ylab('candidates count')


```

The above chart has no surprises as the states with highest population**(Uttar pradesh)** has highest number of candidates. But the second highest in population is **Maharashtra** but it stands at third position in number of candidates after **Bihar**. Also **West bengal** having fourth populated state stands at 9th position in number of candidates.

The data may look surprising but not too much for an Indian who is aware of political system of the conuntry.

**Bihar** and **Uttar pradesh** have been for a long time a hub for political revolution. which gives a mojority of polititians to the country.

**West Bengal** had been in a communist based government for almost 40 years out of 70 years of india's history which always believed to have a not so loving attitude toward democracy. this could explain how **West Bengal** ended up having less candidate despite having large population. 

## 4.Distribution of Proportion get by candidates

First lets see the summary of total vote polled.


```{r echo=FALSE,message=FALSE, warning=FALSE, totalvotepoll}

summary(Nat_election$totvotpoll)

```


Now lets see the summary of vote proportion.

```{r echo=FALSE,message=FALSE, warning=FALSE, proportion}

summary(Nat_election$proportion)

```


In the totvopoll we can see that the median is very small which gives us a rough idea that most of the candidate got a very few votes. This is also reflecting in the proportion summary the median proportion is 0.0066 percent which is insanely small as compared to the max value or the mean. More than 75 percent of the candidate didn't even get a 10 percent of vote share.

**Now lets plot this**

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots_vote}
# plot to see the distribution of votes polled
ggplot(Nat_election, aes(x=totvotpoll)) + 
     geom_bar(stat = 'count',fill = '#C51162',color="#C51162") +
     theme(axis.text.x=element_text(angle=90, size=9)) + theme_minimal() + 
     ggtitle("Distribution for votes polled (National)") + xlab("Votes polled")

```

The data is highly skewed as most of the candidates lost due to few no of votes. 
It will be good to see the data in log10 transformation.

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots_vote_log}

# plot to see the distribution of votes shared
ggplot(Nat_election, aes(x=totvotpoll)) + 
     geom_bar(stat = 'count',fill = '#C51162',color="#C51162") +
     theme(axis.text.x=element_text(angle=90, size=9)) + scale_x_log10() + theme_minimal() + 
     ggtitle("Distribution for votes polled (National)") + xlab("Votes polled(log10)")

```

log10 transformation gives a better picture of data. We almost got a normal distribution.

## 5. Which state loves women more
```{r echo=FALSE,warning=FALSE, univariate_Plots_5}

#filtering out the winning candidates
state_women_won = subset(Nat_election,won == TRUE & cand_sex == 'F') 

#plotting winning candidates by states
ggplot(state_women_won, aes(x=st_name)) + 
  geom_bar(stat = 'count',fill = '#C51162',color="white") +
  geom_text(stat="count", color="black", vjust = 0.4,hjust=-0.2, size=3, aes(y=..count.., label=..count..)) + coord_flip()+
  theme(axis.text.x=element_text(angle=90, size=9)) +
   theme_minimal() +
  ggtitle("Women Won From states") + ylab('contestant count')

```

**Uttar pradesh** is at the top where maximum womens have won in the election, followed by **westbengal** and **Bihar**. There are Many states in which women politicians rarely won like **tripur**, **sikkim**, **jharkhand** etc.

## 6. Cast Divison In Parliament

India is a cast Based society, It thousands of casts which play an important role in election, there are some reserved seats in the parliament on which candidates from a particular cast can contest. Lets see how many cast reserve seats have been in the parliament through out the years of election.

There are two main cast who has reservation 

**1)** Schedule cast(SC)
**2)** Schedule cast(ST)

Every body else contest in **genral** category.

```{r echo=FALSE,warning=FALSE, univariate_Plots_6}

#filtering out the winning candidates and excluding candidate with no cast type
won_candidates = subset(Nat_election,won == TRUE & pc_type != '') 

#little bit more cleaning
won_candidates$pc_type[won_candidates$pc_type == 'SC '] <- 'SC'
won_candidates$pc_type <- factor(won_candidates$pc_type) 

#summary of the winning candidates by cast type
summary(won_candidates$pc_type)

#plotting cast vs count through out the yars
ggplot(won_candidates, aes(x=pc_type)) + 
  geom_bar(stat = 'count',fill = '#F44336',color="white") +
  geom_text(stat="count", color="white", vjust = 0.4,hjust=1.1, size=3, aes(y=..count.., label=..count..)) + coord_flip()+
  theme(axis.text.x=element_text(angle=90, size=9)) +
   theme_minimal() + facet_wrap(~year) +
  ggtitle("Reserve Seats for different casts") + ylab('contestant count') +xlab("Casts")

```

We can see from above that the seats reserves for candidates from particular cast are barely increased. for **SC** it is from **78** in **1977** to **84** in **2014** and for **ST** it went from **38** in **1977** to **47** in **2014**.

This may reflect that cast dominance is incresed in India through out these years.  

# Univariate Analysis

### Structure

The structure of National election data is simple. As seen at the start of the analysis.
each row represents a candidate contested for a particular year.
we have **11** columns:

* st_name (Name of state)
* year (year of election)
* pc_no (parliamentry constituency no)
* pc_name (parliamentry constituency name)
* pc_type (parliamentry constituency type)
* cand_name (candidate name)
* cand_sex (candidate sex)
* partyname (political party name)
* partyabbre (party's short form)
* totvotpoll (total votes candidate get)
* electors (total electors present in the constituency)

st_name, pc_name, pc_type, cand_name, cand_sex, partyname and partyabbre are of type **factor** and year, pc_no, totvotpoll and electors are of type **integer**.

### Main feature(s) of interest

party_name, st_name, year, cand_sex are essential feature of intrest which allowed me to answes much fundamental questions like-

* Which state has most electors?
* Does women in India has the capacity to run for election?
* Which year we see the most candidate run for election?

and many more question.

### Other features which will help support my investigation into my feature(s) of interest?

totvotpoll, electors are the other two main feature which will help me explore few more relation ship like-

* How vote share differs among gender?
* How vote share differ based on electors?
* Are independent candidates able to attract votes through out the years?
* How states differ in vote shares?

### Did you create any new variables from existing variables in the dataset?
I made three new variable -

* **proportion** - how much portion of the total electors a candidate get.
* **win_vote** - The vote share of the winning cadidate for that constituency in that year.
* **won** - represents a particular candidate won the election or not.

### Unusual distributions and its remedy

There are constituencies whose win_vote is 0. which cannot be true. The reason might be that election could be halted for that constituency by the court or election comminsion, which happened manier times in india. So I removed those entries as a outliar.

### Observations

* Among all the election 1996 is the year where maximum candidates contested the election.
* Other gender and female candidates are way fewer than the male candidates.
* Most of the candidates are from **Uttar Pradesh**(most populated state). But **west Bengal**(4th populated state) after having a huge population stands 7th in the no of candidate contested.
* Most of Winning women candidate are from **Uttar Pradesh** followed by **west Bengal**.
* Seats reserved for SC and ST has increased through out the years.

# Bivariate Plots and analysis

Now we are going to dig a little deeper in our analysis by adding one more dimension based on what we saw intresting in above plots. 

## 1. Proportion based on gender

```{r echo=FALSE,warning=FALSE, Bivariate_Plots}

#removing the observation with null values
Nat_election_gender <- subset(Nat_election,cand_sex != 'NULL')
Nat_election_gender$cand_sex <- factor(Nat_election_gender$cand_sex)

#summary total votes by gender
by(Nat_election_gender$proportion, Nat_election_gender$cand_sex, summary)

#plotting the proportion vs gender plot
ggplot(Nat_election_gender, aes(x=cand_sex, y=proportion, fill = cand_sex)) + geom_boxplot() +
  scale_y_continuous(limits = c(0,0.01),breaks=seq(0, 0.01, .002)) +
  ggtitle('Proportion Boxplot By gender')


```

From the above Boxplot and summary we can say that most women in percentage has little bit better proportion of votes than men. Though women candidates are few, We have to see the female contribution as compared to men a little bit closer.

## 2.Women candidate By state
```{r echo=FALSE,warning=FALSE, Bivariate_Plots_1}

#Density plot winning and loosing candidate based on gender
ggplot(subset(Nat_election,cand_sex != 'O' & cand_sex != '' & cand_sex != 'NULL'), aes(x=proportion, fill=won)) +
  geom_density(binwidth=10, alpha=0.5) + scale_y_continuous(limits = c(0,10)) + facet_wrap(~cand_sex) +
  ggtitle('Win/loose vs vote proportion Density chart')


```

We can see that both men and women who won the election share same distribution of proportion.

## 3. Women candidate By party

Lets see which party fielded more women candidate through out the years and how being in a  party effect the winning of a women candidate.
```{r echo=FALSE,warning=FALSE, Bivariate_Plots_2}

#Grouping the Females candidates By party and their status
Nat_election_party_women <- Nat_election %>%
          filter(cand_sex == 'F') %>%
          group_by(partyabbre,won) %>%
          summarise(count_no = n())

#sortion the data by no. of candidates by each party
Nat_election_party_women_sorted = Nat_election_party_women[order(-Nat_election_party_women$count_no),]

#Plotting Female contestant vs Political Party
P1 <- ggplot(head(subset(Nat_election_party_women_sorted,won == FALSE),10), aes(x=reorder(partyabbre,count_no),y=count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') +
  geom_text(stat="identity", color="white", vjust = 0.35, hjust=1, size=3, aes(y=count_no, label=count_no)) +
  theme(axis.text.x=element_text(angle=90, size=9)) +
  coord_flip() + theme_minimal() +
  ggtitle("No. of Female Cadidates by political parties who looses") + 
  xlab("Political Parties") + ylab("Female candidates count")

P2 <- ggplot(head(subset(Nat_election_party_women_sorted,won == TRUE),10), aes(x=reorder(partyabbre,count_no),y=count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') +
  geom_text(stat="identity", color="white", vjust = 0.35, hjust=1.5, size=3, aes(y=count_no, label=count_no)) +
  theme(axis.text.x=element_text(angle=90, size=9)) +
  coord_flip() + theme_minimal() + scale_y_continuous(limits = c(0,200))+
  ggtitle("No. of Female Cadidates by political parties who Won") + 
  xlab("Political Parties") + ylab("Female candidates count")

grid.arrange(P1, P2)

```

Here we can see that among top 10, the most of the womens contested as **Independent** and lost, It is very Hard for a women to contest and win an election as independent. But on the other hand if contesting women is in a political party, her chances to win is better.

**Indian National Congress** is at the top most political party whoose winning women candiate is higher than any other party followed by **Bhartiya Janta Party**.

## 3. Male candidate By party

Lets see which party fielded more women candidate through out the years and how being in a  party effect the winning of a women candidate.
```{r echo=FALSE,warning=FALSE, Bivariate_Plots_3}

#Grouping the Male candidates By party and their status
Nat_election_party_men <- Nat_election %>%
          filter(cand_sex == 'M') %>%
          group_by(partyabbre,won) %>%
          summarise(count_no = n())

#sortion the data by no. of candidates by each party
Nat_election_party_men_sorted = Nat_election_party_men[order(-Nat_election_party_men$count_no),]

#PLotting Male Contestant vs State
P1 <- ggplot(head(subset(Nat_election_party_men_sorted,won == FALSE),10), aes(x=reorder(partyabbre,count_no),y=count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') +
  geom_text(stat="identity", color="black", vjust = 0.35, hjust=0, size=3, aes(y=count_no, label=count_no)) +
  theme(axis.text.x=element_text(angle=90, size=9)) + scale_y_continuous(limits = c(0,43000)) +
  coord_flip() + theme_minimal() +
  ggtitle("No. of Male Cadidates by political parties who looses") + 
  xlab("Political Parties") + ylab("Male candidates count")

P2 <- ggplot(head(subset(Nat_election_party_men_sorted,won == TRUE),10), aes(x=reorder(partyabbre,count_no),y=count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') + 
  geom_text(stat="identity", color="black", vjust = 0.35, hjust=0, size=3, aes(y=count_no, label=count_no)) +
  theme(axis.text.x=element_text(angle=90, size=9)) + scale_y_continuous(limits = c(0,1700)) +
  coord_flip() + theme_minimal() +
  ggtitle("No. of Male Cadidates by political parties who Won") + 
  xlab("Political Parties") + ylab("Male candidates count")

grid.arrange(P1, P2)

```

From the above data we can say that not just female, Infact male candidates are also vulnerable to faliure if they contested election as Independent. And the cances of winning is more if candiates belongs to a palititcal party.

## 4.Rise And Fall Of Women.

Here we are going to see how many female cadidate make it to the parliament in the election.


```{r echo=FALSE,warning=FALSE, Bivariate_Plots_4}

#filtering out the winning women candidates
Women_won <- subset(Nat_election, cand_sex == 'F' & won == TRUE)

ggplot(Women_won, aes(x=year)) + 
  geom_bar(stat = 'count',fill = '#C51162',color="black") +
  geom_text(stat="count", color="black", vjust = -0.1, size=3, aes(y=..count.., label=..count..)) + 
  theme(axis.text.x=element_text(angle=90, size=9)) +
   theme_minimal() +
  xlab('Election Years') + ylab('candidates count') +
  ggtitle("winning Women candidates by year")

```

As happy to see a slow but yet progressive trend for women taking more seats in parliament through out the years.

## 5.Rise and fall of women's charisma

Here we are going to see how women candiate's vote proportion varies through out the years.

```{r echo=FALSE,warning=FALSE, Bivariate_Plots_5}

#plotting the density line chart of women vote share though out the years
ggplot(Women_won, aes(x=proportion,color=year)) + geom_density(position="dodge") + ggtitle('Vote proportion density chart for women candidates')

```

2014 may be the year when maximum no of womens won the election but their vote share decreases than the previous elections. 

## 6. Respond to Independent's Call

Here we see the the performance of independent candidate through out years.

```{r echo=FALSE,warning=FALSE, Bivariate_Plots_ind}

#getting the independent candidates
independent_cand <- subset(Nat_election,partyabbre == 'IND' & won == TRUE)

ggplot(independent_cand, aes(x=year)) + 
  geom_bar(stat = 'count',fill = '#C51162',color="black") +
  geom_text(stat="count", color="black", vjust = -0.1, size=3, aes(y=..count.., label=..count..)) + 
  theme(axis.text.x=element_text(angle=90, size=9)) +
   theme_minimal() +
  xlab('Election Years') + ylab('candidates count') +
  ggtitle("winning Independent candidates by year")

```

According to the above graph 1991 and 2014 were the worst years for Independent candidates. It is followed by the 2004 election where only 5 candidates won.

To conclude this we can say that Elections in india is a tough job for an Independent candidate rather than a candidate with a political party in his/her back.

## 7. States concious of democracy

Here we see the votepolled out of total electorats in each state.

```{r echo=FALSE,warning=FALSE, Bivariate_Plots_6}

#summing the  votes and electors by grouping them by state
state_grouped <- Nat_election %>%
  group_by(st_name) %>% 
  summarise(polled_vote = sum(totvotpoll),
            total_vote = sum(unique(electors)))

#making a proportion variable
state_grouped$proportion <- state_grouped$polled_vote/state_grouped$total_vote

#plotting the voting percentage by states 
ggplot(state_grouped, aes(x=reorder(st_name,proportion),y=proportion,width=0.8)) + 
  geom_histogram(stat = 'identity',fill = '#C51162',color='black') + 
  scale_y_continuous(labels = percent, limits = c(0,1)) + 
  geom_text(stat="identity", color="black", vjust = 0.35, hjust=-0.2, size=3,aes(y=proportion, label=paste(round(proportion, 3)*100, "%", sep=""))) + 
  theme(axis.text.x=element_text(angle=90, size=9)) +
  coord_flip() + theme_minimal() +
  ggtitle("People who voted in each state") + 
  xlab("States") + ylab("Percentage count of polled votes")

```


Above Bar Graph shows that Lakshdweep an island which is not a state but a union territory has the people which are more concious about election than any other state.

One intresting things in the above chart is that states like **West bengal**, **Nagaland**, **Tripura** which are suffering from gross insurgent military groups in their region has the high vote percentage than other states.

**Kerela** which is the most literate state in the country is also in the top 10.

## 8. Proportion vs Electors

we are going how much a candidate gathere vote based on the electors size in his constiuency.

```{r echo=FALSE,warning=FALSE, Bivariate_Plots_7}

#Filtering out the winning candidates
won_candidate_data <- subset(Nat_election,won == TRUE)

ggplot(won_candidate_data, aes(x=proportion,y=electors)) + geom_jitter(alpha=1/10) +
  scale_x_continuous(labels=percent,limits = c(0,0.8),breaks = seq(0,0.8,0.1)) +
  scale_y_continuous(labels = comma) + xlab("votes percentage") + ylab("total electors") + ggtitle("electors vs Vote proportion of winning candidates")

```

In above scatter plot we can see that most of the candidates won at a prcentage range of 15 to 45 there are very few winning candidate with less than 10% of vote share, also there are very few candidates with vote share grater than 50%. 

One Intresting point we can make here is that there is hardly any candidate having a total electors of more than 200,000 and having a vote share of more than 25% or 30%.

Lets look it in different years.


```{r echo=FALSE,warning=FALSE, Bivariate_Plots_8}

ggplot(won_candidate_data, aes(x=proportion,y=electors)) + geom_jitter(alpha=1/10) +
  scale_x_continuous(labels=percent,limits = c(0,0.8),breaks = seq(0,0.8,0.2)) +
  scale_y_continuous(labels = comma) + xlab("votes percentage") + ylab("total electors") + ggtitle("electors vs Vote proportion of winning candidates") + facet_wrap(~year)

```

```{r echo=FALSE,warning=FALSE, Bivariate_Plots_9}

P1 <- ggplot(subset(won_candidate_data,year == '1977'), aes(x=proportion,y=electors)) + geom_jitter(alpha=1/10) +
  scale_x_continuous(labels=percent,limits = c(0,0.8),breaks = seq(0,0.8,0.2)) +
  scale_y_continuous(labels = comma) + xlab("votes percentage") + ylab("total electors")

P2 <- ggplot(subset(won_candidate_data,year == '2014'), aes(x=proportion,y=electors)) + geom_jitter(alpha=1/10) +
  scale_x_continuous(labels=percent,limits = c(0,0.8),breaks = seq(0,0.8,0.2)) +
  scale_y_continuous(labels = comma) + xlab("votes percentage")

grid.arrange(P1,P2,ncol=2,top ="electors vs Vote share(%) winning candidates")

```
 
 In the First scatterplot(facet by years) we can see a pattern, more are more candidates in every metioned year got a similar vote share but at a much higher total electors.
 
 The very shift in the total electors we can see in the second graph which is of the first election(1977) and the last(2014). Also the variance in 1977 election is greater than in 2014.
 
 There could be two reasons for this:
 
 **1.** Due to population growth.
 
 **2.** Due to a consistent use of politics of populism.
 
 We will investigate furthur on our second point.

# Bivariant Analysis

### Features relations observation and Surprices With that

* Though the women's candidate are lower in numbers but womens secured a good meadia in proportion then men, more women (in %) get good proportion of vote then men.

* There is a slow increase in women winning candidate through out the year but surprizingly their vote share drops down in the later elections.

* Independent candidates are may be the highest in contesting the elecctions but they are most likely ones to loose the election. On the other hand only few political parties gets the majority of seats. 

* The top states which are concious of democracy are either union teritories or the states which are dealing with some kind of a militant insurgency in their state.  


### Strongest Relationship

The most strongest relation is between the electors and the vote percentage. Most of the winning candidate have the vote share between 20% to 40% but it is less likely to have a higher vote share when total electors are very high. 

Also it is very less likely to have a vote share of more than 50% even for a small value of total electors.

Another intresting point in the above relation is that as every next election the electors increses but the variance in the proportion gets slightly decreased. 



# Multivariate Plots Section

## 1. Electors vs proportion by year

```{r echo=FALSE,warning=FALSE, Multivariate_Plots}

#plooting the chart between Electors and vote polled percentage
ggplot(won_candidate_data, aes(x=proportion,y=electors,color=year))+ geom_jitter() + scale_color_brewer(type = 'div')+
  scale_x_continuous(labels=percent,limits = c(0,0.8),breaks = seq(0,0.8,0.1)) +
  scale_y_continuous(labels = comma,limits = c(0,3000000),breaks = seq(0,3000000,500000)) + xlab("votes percentage") + ylab("total electors") + ggtitle("electors vs Vote proportion of winning candidates in years")

```

Here the change in color as we go down to up in total electors represents the population growth through out the years, but also we are looking at a decrease in variance as we grow in poulation.

## 2.Electors vs Vote share By Gender

```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots_2}

#filtering out the women who won the election
women_won <- subset(won_candidate_data,cand_sex == 'F')

#plooting the chart between Electors and vote polled percentage
ggplot(won_candidate_data, aes(x=electors,y=proportion,color=cand_sex))+ geom_smooth(se=FALSE) +
  scale_y_continuous(labels=percent,limits = c(0,0.8),breaks = seq(0,0.8,0.1)) +
  scale_x_continuous(labels = comma,breaks = seq(0,4000000,500000)) + ylab("votes percentage") + xlab("total electors") + ggtitle("electors vs Vote proportion line chart based on gender")

ggplot(won_candidate_data, aes(x=electors,y=proportion,color=year))+ geom_smooth(se=FALSE) +
  scale_y_continuous(labels=percent,limits = c(0,0.8),breaks = seq(0,0.8,0.1)) +
  scale_x_continuous(labels = comma) + xlab("votes percentage") + ylab("total electors") + ggtitle("electors vs Vote proportion line chart of women candidate based on years") 

```

In the first chart we can see that women did better in getting higher votes in constituencies with large electors then men.

In the seconds graph we can see that 2014 has been one of the most successfull year for women candidates for larger constituencues in getting higher vote share.

## 3.Top Political Parties

Now we are going to see how the vote share varies in top political parties through out the year.
```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots_3}

#grouping the winning candidates by party
Nat_election_party <- won_candidate_data %>%
          group_by(partyabbre) %>%
          summarise(cand_count = n())

#soritng the party by candidates
Nat_election_party <- Nat_election_party[order(-Nat_election_party$cand_count),]

#extracting the top 5 party
top_party_name <- head(Nat_election_party,10)

#getting full top parties data
top_parties <- subset(won_candidate_data,partyabbre %in%  top_party_name$partyabbre)
top_parties$partyabbre <- factor(top_parties$partyabbre)

#Plotting top parties vs no of candidates 
ggplot(top_parties, aes(x=partyabbre)) + 
  geom_histogram(stat = 'count',fill = '#C51162') +
  geom_text(stat="count", color="white", vjust = 0.35, hjust=1.2, size=5, aes(y=..count.., label=..count..)) +
  theme(axis.text.x=element_text(angle=90, size=9)) +
  coord_flip() + theme_minimal() +
  ggtitle("Top Parties Winning the elections") + 
  xlab("Political Parties") + ylab("candidates count")


```


We can see that out of top ten parties two parties have the largest share which are the **BJP** and **INC**.

```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots_4}
#getting summary of parties by years
by(top_parties$partyabbre,top_parties$year,summary)

#plooting stacked bar graph with top parties by year
ggplot(data = top_parties, aes(x = year, fill = partyabbre)) + 
    geom_bar(stat = 'count',color="black") + 
  coord_flip() + xlab('Years') + ylab("Parliament seats won by political parties") + 
  ggtitle('Rise and Fall of top parties')
```

From the above summary we can say that INC(Indian National congress) has been the most dominating party through out the years, But we see a fascinating shift in 2014 where INC went down to 44 seats from 206 seats in 2009. And BJP which starts from 2 seats in 1984 went up to 282 seats in 2014.

But Only few political parties get the majority of seats in the election.

Note: 

* **BLD** is an outcome of a grand allaince(merger of many parties) to defeat **INC**. They won the election but the party as well their government didn't last. 

* **INC(I)** was a new party but with the old politicians of **INC**. because the court banned **INC** to contest election, so they made a new party to contest the election.

## 4. Independents Vote share analysis

```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots_5}
#getting The independent candidates who won
independent_won <- subset(won_candidate_data, partyabbre == 'IND')

#womens growth through out election years
ggplot(data = independent_won, aes(x=proportion, y=electors,color=cand_sex)) + geom_jitter() +  scale_x_continuous(labels=percent,limits = c(0,0.8),breaks = seq(0,0.8,0.2)) +
  scale_y_continuous(labels = comma,limits = c(0,2000000),breaks = seq(0,2000000,500000)) +
 xlab('vote share (%)') + ylab("Electors") + 
  ggtitle('Independent candidate vote share vs total electors') + facet_wrap(~year)
```

Independent candidates had experienced a lot of ups and downs through out the years. For women it looks a lost battle. And even for men it's not so good.  

According to above plot its hard to find any pattern in it.

## 5.Top Parties Performance

In this we are going to look at the performance of top three parties which we found above. They are:

* India National Congress (INC)
* BHartiya Janta Party (BJP)
* Communist Party Of India-Marxist (CPM)

```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots_6}

#filtering the top parties
Top_Parties <- subset(won_candidate_data, partyabbre == 'INC' | partyabbre == 'BJP' | partyabbre == 'CPM')

#plotting scatter plot
ggplot(data = Top_Parties, aes(x=proportion, y=electors,color=partyabbre)) + geom_jitter(alpha = 1/2) + 
 xlab('vote share (%)') + ylab("Electors") + scale_x_continuous(labels = percent,breaks = seq(0,1,0.1)) +
  scale_y_continuous(labels = comma,breaks = seq(0,3000000,500000)) +
  ggtitle('Proformance of top political parties')

```

INC may be the longest running party to woo the voters but from the above graph we can say that BJP has an edge in higher electors constituncies.

```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots_7}
#filtering the top parties
INC <- subset(won_candidate_data, partyabbre == 'INC')
BJP <- subset(won_candidate_data, partyabbre == 'BJP')

#plotting performance of INC
ggplot(data = INC, aes(x=proportion, y=electors)) + geom_jitter(alpha = 1/2) + 
 xlab('vote share (%)') + ylab("Electors") + scale_x_continuous(labels = percent,breaks = seq(0,1,0.1)) +
  scale_y_continuous(labels = comma,breaks = seq(0,3000000,500000)) +
  ggtitle('Proformance of INC') + facet_wrap(~year)

#plotting performance of BJP
ggplot(data = BJP, aes(x=proportion, y=electors)) + geom_jitter(alpha = 1/2) + 
 xlab('vote share (%)') + ylab("Electors") + scale_x_continuous(labels = percent,breaks = seq(0,1,0.1)) +
  scale_y_continuous(labels = comma,breaks = seq(0,3000000,500000)) +
  ggtitle('Proformance of BJP') + facet_wrap(~year)

```

Here we are looking at the performance of top two political parties having a very opposite performance chart, One is BJP who went from nothing to majority in the parliment, and the Other is INC who went from Allmost all the seats to just 44 seats in the parliament.

## Party winning candidates through out the years by gender

```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots_8}

#plotting the bar garph of INC by gender
ggplot(INC, aes(x= year, fill=cand_sex) ) +
  geom_histogram(stat="count", position="dodge", color="black") +
  geom_text(stat='count',data=subset(INC, cand_sex =="M"), color="white",
            aes(y=..count.., label=..count..), 
             size=3, vjust = -0.5, hjust=1.3) +
  geom_text(stat='count',data=subset(INC, cand_sex =="F"),color="black",
            aes(y=..count.., label=..count..), 
             size=3, vjust = 1.5, hjust=-0.3) +
  theme_minimal() +
  coord_flip() + 
  scale_fill_manual(values=c("goldenrod1","dodgerblue4" ), name="candidate Sex") + 
  ggtitle("Candidates by gender in INC") + xlab("Years") + ylab("Candidates")

#plotting the bar garph of INC by gender
ggplot(BJP, aes(x= year, fill=cand_sex) ) +
  geom_histogram(stat="count", position="dodge", color="black") +
  geom_text(stat='count',data=subset(BJP, cand_sex =="M"), color="white",
            aes(y=..count.., label=..count..), 
             size=3, vjust = -0.5, hjust=1.3) +
  geom_text(stat='count',data=subset(BJP, cand_sex =="F"),color="black",
            aes(y=..count.., label=..count..), 
             size=3, vjust = 1.5, hjust=-0.3) +
  theme_minimal() +
  coord_flip() + 
  scale_fill_manual(values=c("goldenrod1","dodgerblue4" ), name="candidate Sex") + 
  ggtitle("Candidates by gender in BJP") + xlab("Years") + ylab("Candidates")

```

FOr both the parties women share in each year is always less than 15 percent and in most of the cases its less than 10%. but we see a overall increase in women cadidates winnning in the recent years. 

For BJP 2014 is the first time that 30 women(highest ever in BJP) won the election.


# Final Plots and Summary

### 1. Cast Divison In Parliament

India is a cast Based society, It thousands of casts which play an important role in election, there are some reserved seats in the parliament on which candidates from a particular cast can contest. Lets see how many cast reserve seats have been in the parliament through out the years of election.

There are two main cast who has reservation 

**1)** Schedule cast(SC)
**2)** Schedule cast(ST)

Every body else contest in **genral** category.
```{r echo=FALSE,warning=FALSE, final_pot_1}

#plotting cast vs count through out the yars
ggplot(won_candidates, aes(x=pc_type)) + 
  geom_bar(stat = 'count',fill = '#F44336',color="white") +
  geom_text(stat="count", color="white", vjust = 0.4,hjust=1.1, size=3, aes(y=..count.., label=..count..)) + coord_flip()+
  theme(axis.text.x=element_text(angle=90, size=9)) +
   theme_minimal() + facet_wrap(~year) +
  ggtitle("Reserve Seats for different casts") + ylab('contestant count') +xlab("Casts")

```

We can see from above that the seats reserves for candidates from particular cast were increased. for **SC** it is from **78** in **1977** to **84** in **2014** and for **ST** it went from **38** in **1977** to **47** in **2014**. As a result of that seats comes under general category gets decreased.

This reflect how cast dominance is incresed in India through out these years.

### 2. Women/Men candidate By party/Independent

```{r echo=FALSE,warning=FALSE, final_plot_2}

#Plotting Female contestant vs Political Party
P1 <- ggplot(head(subset(Nat_election_party_women_sorted,won == FALSE),10), aes(x=reorder(partyabbre,count_no),y=count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') +
  geom_text(stat="identity", color="white", vjust = 0.35, hjust=1, size=3, aes(y=count_no, label=count_no)) +
  theme(axis.text.x=element_text(angle=90, size=9)) +
  coord_flip() + theme_minimal() +
  ggtitle("No. of Female Cadidates by political parties who looses") + 
  xlab("Political Parties") + ylab("Female candidates count")

P2 <- ggplot(head(subset(Nat_election_party_women_sorted,won == TRUE),10), aes(x=reorder(partyabbre,count_no),y=count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') +
  geom_text(stat="identity", color="white", vjust = 0.35, hjust=1.5, size=3, aes(y=count_no, label=count_no)) +
  theme(axis.text.x=element_text(angle=90, size=9)) +
  coord_flip() + theme_minimal() + scale_y_continuous(limits = c(0,200))+
  ggtitle("No. of Female Cadidates by political parties who Won") + 
  xlab("Political Parties") + ylab("Female candidates count")

grid.arrange(P1, P2)

P3 <- ggplot(head(subset(Nat_election_party_men_sorted,won == FALSE),10), aes(x=reorder(partyabbre,count_no),y=count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') +
  geom_text(stat="identity", color="black", vjust = 0.35, hjust=0, size=3, aes(y=count_no, label=count_no)) +
  theme(axis.text.x=element_text(angle=90, size=9)) + scale_y_continuous(limits = c(0,43000)) +
  coord_flip() + theme_minimal() +
  ggtitle("No. of Male Cadidates by political parties who looses") + 
  xlab("Political Parties") + ylab("Male candidates count")

P4 <- ggplot(head(subset(Nat_election_party_men_sorted,won == TRUE),10), aes(x=reorder(partyabbre,count_no),y=count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') + 
  geom_text(stat="identity", color="black", vjust = 0.35, hjust=0, size=3, aes(y=count_no, label=count_no)) +
  theme(axis.text.x=element_text(angle=90, size=9)) + scale_y_continuous(limits = c(0,1700)) +
  coord_flip() + theme_minimal() +
  ggtitle("No. of Male Cadidates by political parties who Won") + 
  xlab("Political Parties") + ylab("Male candidates count")

grid.arrange(P3, P4)

```

Here we can see that among top 10, the most of the women/men contested as **Independent**(IND) and lost, It is very Hard for any candidate to contest and win an election as independent. But on the other hand if contesting women/men is in a political party, his/her chances to win is better.

One Big fact which is revealed by the above chart is how much less women were fielded in the election as compared to men. This shows a gender bias in among all the political parities. 

**Indian National Congress**(INC) is at the top most political party whoose winning women candiate is higher than any other party followed by **Bhartiya Janta Party**(BJP).


### 3.Rise And Fall Of Women.

Here we are going to see how many female cadidate make it to the parliament in the election.

```{r echo=FALSE,warning=FALSE, final_plot_3}

ggplot(Women_won, aes(x=year)) + 
  geom_bar(stat = 'count',fill = '#C51162',color="black") +
  geom_text(stat="count", color="black", vjust = -0.1, size=3, aes(y=..count.., label=..count..)) + 
  theme(axis.text.x=element_text(angle=90, size=9)) +
   theme_minimal() +
  xlab('Election Years') + ylab('candidates count') +
  ggtitle("winning Women candidates by year")

```

After having a huge gender bias among political parties, **Women didn't gave up**, this showed by above chart, which shows a slow but a hopefull increase in women in the parliament thorugh out these years. 

### 4. States concious of democracy

We will see in this graph how much citizen of each state take election seriously or went to poll their votes.

```{r echo=FALSE,warning=FALSE, final_plot_4}

#plotting the voting percentage by states 
ggplot(state_grouped, aes(x=reorder(st_name,proportion),y=proportion,width=0.8)) + 
  geom_histogram(stat = 'identity',fill = '#C51162',color='black') + 
  scale_y_continuous(labels = percent, limits = c(0,1)) + 
  geom_text(stat="identity", color="black", vjust = 0.35, hjust=-0.2, size=3,aes(y=proportion, label=paste(round(proportion, 3)*100, "%", sep=""))) + 
  theme(axis.text.x=element_text(angle=90, size=9)) +
  coord_flip() + theme_minimal() +
  ggtitle("People who voted in each state") + 
  xlab("States") + ylab("Percentage count of polled votes")

```


Above Bar Graph shows that Lakshdweep an island which is not a state but a union territory has the population which are more concious about election than any other state.

One intresting things in the above chart is that states like **West bengal**, **Nagaland**, **Tripura** which are suffering from gross insurgent military groups in their region has the high vote percentage than other states.

**Kerela** which is the most literate state in the country is also in the top 10.

### 5. Electors vs proportion by year

```{r echo=FALSE,warning=FALSE, final_plot_5}

#plooting the chart between Electors and vote polled percentage
ggplot(won_candidate_data, aes(x=proportion,y=electors,color=year))+ geom_jitter() + 
  scale_color_brewer(type = 'div') +
  scale_x_continuous(labels=percent,limits = c(0,0.8),breaks = seq(0,0.8,0.1)) +
  scale_y_continuous(labels = comma,limits = c(0,3000000),breaks = seq(0,3000000,500000)) + 
  xlab("votes percentage") + ylab("total electors") + 
  ggtitle("electors vs Vote proportion of winning candidates in years")

```

The above chart represents how rapidly the constituencies population had increased through out the years. Also how the proportion of vote remain almost same for the winning candidates. Most of the winning candidate get the vote share between 10% and 50%.

But as we go up we see a decrease in the variance of voteshare, like in 2014, the variance in the vote share decreased, most of the candidates seems to get vote share between 15% to 45%.

### 6. Political Shift in Parliament

```{r echo=FALSE,message=FALSE, warning=FALSE, final_plot_6}

#plooting stacked bar graph with top parties by year
ggplot(data = top_parties, aes(x = year, fill = partyabbre)) + 
    geom_bar(stat = 'count',color="black") + 
  coord_flip() + xlab('Years') + ylab("Parliament seats won by political parties") + 
  ggtitle('Rise and Fall of top parties')

```

From the above chart we can say that **INC(Indian National congress)** has been the most dominating party through out the years, but we see a fascinating shift in 2014 where INC went down to 44 seats from 206 seats in 2009. And BJP which starts from 2 seats in 1984 went up to 282 seats in 2014.

But the most bitter truth of all is that only few political parities took majority of seats.

Note: 

* **BLD** is an outcome of a grand allaince(merger of many parties) to defeat **INC**. They won the election but the party as well their government didn't last. 

* **INC(I)** which won in 1980 was a new party but with the old politicians of **INC**. because the court banned **INC** to contest election, so they made a new party to contest the election.

------

# Reflection

### Issues with the data

 The data has several issues.

  * There is very few information given about the candidate.
  
  * The data is not so dirty but I had to remove some observation as it gets zero votes and also won the election.
  
  * I had to make few variables like won_vote(represents the vote of winning cadidate), won(represent weather a candidate won or not), proportion(how much percentage of vote share a cadidate get)

  * Many political party had party names with different spellings so I had to use party's abbreviation.

  * I had to change many columns into factors.

### Achieved in getting
 
  * A cleaner picture of parlimentary seats reservtion.
 
  * A rough idea how winning cadidates proportion related to voters population in their constituencies.
 
  * How dominance of top political parties shift over the years.
 
  * See a clear gender bias among the political parties.
 
  * Performance of women, performance of independent candidates.
 
  * get a fair picture about which state's cititzer are intrested/concious about election or voting.
 
  * How only few political parties ruling India for so long.

### Could be done in future
 
  * performance of winning candidates who run for election more than one time.
  
  * performance of runnerups who run for next election.
  
  * performance of political parties based on how many candidate they change in every election.
  
  * which political party is dominating which state.
  
### Refrences

  * https://www.rdocumentation.org/packages/data.table/versions/1.10.4-2/topics/merge
  
  * http://ggplot2.tidyverse.org/reference/coord_flip.html
  
  * http://ggplot2.tidyverse.org/reference/geom_text.html
  
  * https://en.wikipedia.org/wiki/Indian_general_election,_2014
  
  * https://en.wikipedia.org/wiki/Indian_general_election,_1977
  
  * https://en.wikipedia.org/wiki/Insurgency_in_Northeast_India
  
  * https://en.wikipedia.org/wiki/Elections_in_India
  
  * https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/26526
  
