
Analysis of Elections in India
========================================================

> **Source**:  https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/26526


### Initial Exploration 
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Loading all the required packages
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(gridExtra)
library(extrafont)
library(RColorBrewer)

#setting the working directory
setwd("C:\\Users\\mohdha\\Documents\\GitHub\\India-Election-Analysis-with-R")

```

### Structure of the data

```{r echo=FALSE, Load_the_Data}

# Load the Data
national_election = read.table('nationalelection.tab',sep="\t", header=TRUE)

# looking the structure of data
str(national_election)

```

### Adding new variables

```{r echo=FALSE, making_variables}

#Making a variable proportion to reflect the vote proportion of each candidate
national_election$proportion = national_election$totvotpoll / national_election$electors

head(national_election)

#making a variable with the winning vote
national_election.winner_vote <- national_election %>%
  group_by(st_name, year,pc_no) %>%
  summarise(win_vote = max(totvotpoll))

#making the left join to the orignal data
Nat_election <- 
  merge(national_election,
        national_election.winner_vote,
        by = c('st_name','year','pc_no'),
        x.all = TRUE )


#making our won variable to reflect the winning status of the candidate
Nat_election$won = ifelse(Nat_election$totvotpoll == Nat_election$win_vote, TRUE, FALSE)

summary(Nat_election)


```

### Cleaning the data

```{r echo=FALSE, remove_outliars}
#filtering the outliars
Nat_election <- subset(Nat_election, win_vote != 0)
sprintf('Nation election data has %s Observation after filtering out Outliars.',NROW(Nat_election))

```

```{r echo=FALSE, updated_summary}

#updated summary
summary(Nat_election$win_vote)

```
# Univariate Plots Section

Now We are going to analyse out data in one dimention.

## 1. No of contestant contesting every election year
```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots}

#converting year variable into type factor
Nat_election$year <- as.factor(Nat_election$year)

#plotting the no of candidates contested each year in election
ggplot(Nat_election, aes(x=year)) + 
  geom_histogram(stat = 'count',fill = '#C51162',color="black") +
  theme(axis.text.x=element_text(angle=90, size=9)) +
  coord_flip() + 
  theme_minimal() + 
  scale_y_continuous(breaks=seq(0, 15000, 1000))+
  ggtitle("No of Contestant each year (National)")


```
## 2.Number of contestant based on gender.  

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots_sex}


#plotting the no of candidates contested by gender
ggplot(subset(Nat_election,cand_sex != 'NULL'), aes(x = cand_sex)) + 
  geom_bar(stat = 'count',fill = '#C51162',color = "black") +
  geom_text(stat = "count", 
            color = "black", 
            vjust = -0.1, 
            size = 3, 
            aes(y = ..count.., label = ..count..)) + 
  theme(axis.text.x = element_text(angle = 90, size = 9)) +
   theme_minimal() +
  ggtitle("Contestant by Gender (National)")


```

## 3.Number of contestant based on State.

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots_state}

#plots the histogram to see the no of contestant from each state
ggplot(Nat_election, aes(x = reorder(st_name, st_name, function(x) length(x)))) + 
  geom_histogram(stat = 'count',fill = '#C51162',color = "black") +
  theme(axis.text.x = element_text(angle = 90, size = 9)) +
  coord_flip() + 
  theme_minimal() +
  xlab("States") + 
  ylab('candidates count') +
  ggtitle("No of Contestant from each state (National)")


```


## 4.Distribution of Proportion get by candidates

```{r echo=FALSE,message=FALSE, warning=FALSE, totalvotepoll}

summary(Nat_election$totvotpoll)

```
```{r echo=FALSE,message=FALSE, warning=FALSE, proportion}

summary(Nat_election$proportion)

```

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots_vote}
# plot to see the distribution of votes polled
ggplot(Nat_election, aes(x = totvotpoll)) + 
     geom_bar(stat = 'count',fill = '#C51162',color = "#C51162") +
     theme(axis.text.x = element_text(angle = 90, size = 9)) + 
     theme_minimal() + 
     ggtitle("Distribution for votes polled (National)") + xlab("Votes polled")

```
```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots_vote_log}

# plot to see the distribution of votes shared
ggplot(Nat_election, aes(x = totvotpoll)) + 
     geom_bar(stat = 'count',fill = '#C51162',color = "#C51162") +
     theme(axis.text.x = element_text(angle = 90, size = 9)) + scale_x_log10() + theme_minimal() + 
     ggtitle("Distribution for votes polled (National)") + xlab("Votes polled(log10)")

```



## 5. Which state loves women more
```{r echo=FALSE,warning=FALSE, univariate_Plots_5}

#filtering out the winning candidates
state_women_won = subset(Nat_election,won == TRUE & cand_sex == 'F') 

#plotting winning candidates by states
ggplot(state_women_won, 
       aes(x = reorder(st_name, st_name, function(x) length(x)))) + 
  geom_bar(stat = 'count',fill = '#C51162',color = "white") +
  geom_text(stat = "count", 
            color = "black", 
            vjust = 0.4,
            hjust = -0.2, 
            size = 3,
            aes(y = ..count.., label = ..count..)) + 
  coord_flip()+
  theme(axis.text.x = element_text(angle = 90, size = 9)) +
  theme_minimal() +
  ylab('contestant count') +
  ggtitle("Women Won From states")

```


## 1. Proportion based on gender

```{r echo=FALSE,warning=FALSE, Bivariate_Plots}

#removing the observation with null values
Nat_election_gender <- subset(Nat_election,cand_sex != 'NULL')
Nat_election_gender$cand_sex <- factor(Nat_election_gender$cand_sex)

#summary total votes by gender
by(Nat_election_gender$proportion, Nat_election_gender$cand_sex, summary)

#plotting the proportion vs gender plot
ggplot(Nat_election_gender, 
       aes(x = cand_sex, y = proportion, fill = cand_sex)) + 
  geom_boxplot() +
  scale_y_continuous(limits = c(0,0.01), breaks=seq(0, 0.01, .002)) +
  ggtitle('Proportion Boxplot By gender')


```

## 2.Women candidate By state
```{r echo=FALSE,warning=FALSE, Bivariate_Plots_1}

#Density plot winning and loosing candidate based on gender
ggplot(subset(Nat_election,cand_sex != 'O' & cand_sex != '' & cand_sex != 'NULL'),
       aes(x = proportion, fill = won)) +
  geom_density(binwidth = 10, alpha = 0.5) + 
  scale_y_continuous(limits = c(0,10)) + 
  facet_wrap(~cand_sex) +
  ggtitle('Win/loose vs vote proportion Density chart')


```

## 3. Women candidate By party

```{r echo=FALSE,warning=FALSE, Bivariate_Plots_2}

#Grouping the Females candidates By party and their status
Nat_election_party_women <- Nat_election %>%
          filter(cand_sex == 'F') %>%
          group_by(partyabbre,won) %>%
          summarise(count_no = n())

#sortion the data by no. of candidates by each party
Nat_election_party_women_sorted = Nat_election_party_women[order(-Nat_election_party_women$count_no),]

#Plotting Female contestant vs Political Party
P1 <- ggplot(head(subset(Nat_election_party_women_sorted,won == FALSE),10),
             aes(x = reorder(partyabbre,count_no),y = count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') +
  geom_text(stat = "identity", 
            color = "white", 
            vjust = 0.35, 
            hjust = 1, 
            size = 3, 
            aes(y = count_no, label = count_no)) +
  theme(axis.text.x = element_text(angle = 90, size = 9)) +
  coord_flip() + 
  theme_minimal() +
  xlab("Political Parties") + 
  ylab("Female candidates count") +
  ggtitle("No. of Female Cadidates by political parties who looses")

P2 <- ggplot(head(subset(Nat_election_party_women_sorted,won == TRUE),10),
             aes(x = reorder(partyabbre,count_no),y = count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') +
  geom_text(stat = "identity",
            color = "white",
            vjust = 0.35,
            hjust = 1.5,
            size = 3,
            aes(y = count_no, label = count_no)) +
  theme(axis.text.x = element_text(angle = 90, size = 9)) +
  coord_flip() + 
  theme_minimal() +
  scale_y_continuous(limits = c(0,200)) +
  xlab("Political Parties") + 
  ylab("Female candidates count") +
  ggtitle("No. of Female Cadidates by political parties who Won")

grid.arrange(P1, P2)

```
## 3. Male candidate By party


```{r echo=FALSE,warning=FALSE, Bivariate_Plots_3}

#Grouping the Male candidates By party and their status
Nat_election_party_men <- Nat_election %>%
          filter(cand_sex == 'M') %>%
          group_by(partyabbre,won) %>%
          summarise(count_no = n())

#sortion the data by no. of candidates by each party
Nat_election_party_men_sorted = Nat_election_party_men[order(-Nat_election_party_men$count_no),]

#PLotting Male Contestant vs State
P1 <- ggplot(head(subset(Nat_election_party_men_sorted,won == FALSE),10),
             aes(x = reorder(partyabbre,count_no),y = count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') +
  geom_text(stat = "identity", 
            color =  "black", 
            vjust = 0.35, 
            hjust = 0,
            size = 3,
            aes(y = count_no, label = count_no)) +
  theme(axis.text.x = element_text(angle = 90, size = 9)) + 
  scale_y_continuous(limits = c(0,43000)) +
  coord_flip() + theme_minimal() +
  xlab("Political Parties") + 
  ylab("Male candidates count") +
  ggtitle("No. of Male Cadidates by political parties who looses")

P2 <- ggplot(head(subset(Nat_election_party_men_sorted,won == TRUE),10),
             aes(x = reorder(partyabbre,count_no), y = count_no)) + 
  geom_histogram(stat = 'identity',fill = '#C51162') + 
  geom_text(stat = "identity",
            color = "black",
            vjust = 0.35,
            hjust = 0,
            size = 3,
            aes(y = count_no, label = count_no)) +
  theme(axis.text.x = element_text(angle = 90, size = 9)) + 
  scale_y_continuous(limits = c(0,1700)) +
  coord_flip() + theme_minimal() +
  xlab("Political Parties") +
  ylab("Male candidates count") +
  ggtitle("No. of Male Cadidates by political parties who Won") 
  

grid.arrange(P1, P2)

```


## 4.Rise And Fall Of Women.




```{r echo=FALSE,warning=FALSE, Bivariate_Plots_4}

#filtering out the winning women candidates
Women_won <- subset(Nat_election, cand_sex == 'F' & won == TRUE)

ggplot(Women_won, aes(x = year)) + 
  geom_bar(stat = 'count',fill = '#C51162',color = "black") +
  geom_text(stat = "count",
            color = "black",
            vjust = -0.1,
            size = 3,
            aes(y = ..count.., label = ..count..)) + 
  theme(axis.text.x = element_text(angle = 90, size = 9)) +
  theme_minimal() +
  xlab('Election Years') + 
  ylab('candidates count') +
  ggtitle("winning Women candidates by year")

```


## 5.Rise and fall of women's charisma



```{r echo=FALSE,warning=FALSE, Bivariate_Plots_5}

#plotting the density line chart of women vote share though out the years
ggplot(Women_won, aes(x = proportion,color = year)) + 
  geom_density(position = "dodge") + 
  ggtitle('Vote proportion density chart for women candidates')

```



## 6. Respond to Independent's Call



```{r echo=FALSE,warning=FALSE, Bivariate_Plots_ind}

#getting the independent candidates
independent_cand <- subset(Nat_election,partyabbre == 'IND' & won == TRUE)

ggplot(independent_cand, aes(x = year)) + 
  geom_bar(stat = 'count',fill = '#C51162',color = "black") +
  geom_text(stat = "count",
            color = "black",
            vjust = -0.1,
            size = 3,
            aes(y = ..count.., label = ..count..)) + 
  theme(axis.text.x = element_text(angle = 90, size = 9)) +
  theme_minimal() +
  xlab('Election Years') + 
  ylab('candidates count') +
  ggtitle("winning Independent candidates by year")

```



## 7. States concious of democracy


```{r echo=FALSE,warning=FALSE, Bivariate_Plots_6}

#summing the  votes and electors by grouping them by state
state_grouped <- Nat_election %>%
  group_by(st_name) %>% 
  summarise(polled_vote = sum(totvotpoll),
            total_vote = sum(unique(electors)))

#making a proportion variable
state_grouped$proportion <- state_grouped$polled_vote/state_grouped$total_vote

#plotting the voting percentage by states 
ggplot(state_grouped, 
       aes(x = reorder(st_name,proportion),y = proportion,width=0.8)) + 
  geom_histogram(stat = 'identity',fill = '#C51162',color = 'black') + 
  scale_y_continuous(labels = percent, limits = c(0,1)) + 
  geom_text(stat = "identity",
            color = "black",
            vjust = 0.35,
            hjust = -0.2,
            size = 3,
            aes(y = proportion, 
                label = paste(round(proportion, 3)*100, "%", sep = ""))) + 
  theme(axis.text.x = element_text(angle = 90, size = 9)) +
  coord_flip() + 
  theme_minimal() +
  xlab("States") + 
  ylab("Percentage count of polled votes") +
  ggtitle("People who voted in each state")

```

```{r echo=FALSE,warning=FALSE, Bivariate_Plots_7}

#Filtering out the winning candidates
won_candidate_data <- subset(Nat_election,won == TRUE)

ggplot(won_candidate_data, 
       aes(x = proportion,y = electors)) + geom_jitter(alpha = 1/10) +
  scale_x_continuous(labels = percent,limits = c(0,0.8),breaks = seq(0,0.8,0.1)) +
  scale_y_continuous(labels = comma) + 
  xlab("votes percentage") + 
  ylab("total electors") + 
  ggtitle("electors vs Vote proportion of winning candidates")

```
```{r echo=FALSE,warning=FALSE, Bivariate_Plots_8}

ggplot(won_candidate_data, aes(x = proportion,y = electors)) + 
  geom_jitter(alpha = 1/10) +
  scale_x_continuous(labels = percent,limits = c(0,0.8),breaks = seq(0,0.8,0.2)) +
  scale_y_continuous(labels = comma) + 
  xlab("votes percentage") + 
  ylab("total electors") +
  ggtitle("electors vs Vote proportion of winning candidates") +
  facet_wrap(~year)

```

```{r echo=FALSE,warning=FALSE, Bivariate_Plots_9}

P1 <- ggplot(subset(won_candidate_data,year == '1977'), 
             aes(x = proportion,y = electors)) + 
  geom_jitter(alpha = 1/10) +
  scale_x_continuous(labels = percent,limits = c(0,0.8),breaks = seq(0,0.8,0.2)) +
  scale_y_continuous(labels = comma) +
  xlab("votes percentage") +
  ylab("total electors")

P2 <- ggplot(subset(won_candidate_data,year == '2014'), aes(x = proportion,y = electors)) + 
  geom_jitter(alpha = 1/10) +
  scale_x_continuous(labels = percent,limits = c(0,0.8),breaks = seq(0,0.8,0.2)) +
  scale_y_continuous(labels = comma) + 
  xlab("votes percentage")

grid.arrange(P1,P2,ncol=2,top ="electors vs Vote share(%) winning candidates")

```



# Multivariate Plots Section

## 1. Electors vs proportion by year

```{r echo=FALSE,warning=FALSE, Multivariate_Plots}

#plooting the chart between Electors and vote polled percentage
ggplot(won_candidate_data, 
       aes(x = proportion,y = electors,color = year))+ 
  geom_jitter() + 
  scale_color_brewer(type = 'div')+
  scale_x_continuous(labels = percent,limits = c(0,0.8),breaks = seq(0,0.8,0.1)) +
  scale_y_continuous(labels = comma,limits = c(0,3000000),breaks = seq(0,3000000,500000)) +
  xlab("votes percentage") + 
  ylab("total electors") + 
  ggtitle("electors vs Vote proportion of winning candidates in years")

```

## 2.Electors vs Vote share By Gender

```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots_2}

#filtering out the women who won the election
women_won <- subset(won_candidate_data,cand_sex == 'F')

#plooting the chart between Electors and vote polled percentage
ggplot(won_candidate_data, aes(x = electors,y = proportion,color = cand_sex))+ 
  geom_smooth(se = FALSE) +
  scale_y_continuous(labels = percent,limits = c(0,0.8),breaks = seq(0,0.8,0.1)) +
  scale_x_continuous(labels = comma,breaks = seq(0,4000000,500000)) + 
  ylab("votes percentage") +
  xlab("total electors") + 
  ggtitle("electors vs Vote proportion line chart based on gender")

ggplot(won_candidate_data, aes(x = electors,y = proportion,color = year))+ 
  geom_smooth(se = FALSE) +
  scale_y_continuous(labels = percent,limits = c(0,0.8),breaks = seq(0,0.8,0.1)) +
  scale_x_continuous(labels = comma) +
  xlab("votes percentage") + 
  ylab("total electors") +
  ggtitle("electors vs Vote proportion line chart of women candidate based on years") 

```

In the first chart we can see that women did better in getting higher votes in constituencies with large electors then men.

In the seconds graph we can see that 2014 has been one of the most successfull year for women candidates for larger constituencues in getting higher vote share.

## 3.Top Political Parties

Now we are going to see how the vote share varies in top political parties through out the year.
```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots_3}

#grouping the winning candidates by party
Nat_election_party <- won_candidate_data %>%
          group_by(partyabbre) %>%
          summarise(cand_count = n())

#soritng the party by candidates
Nat_election_party <- Nat_election_party[order(-Nat_election_party$cand_count),]

#extracting the top 5 party
top_party_name <- head(Nat_election_party,10)

#getting full top parties data
top_parties <- subset(won_candidate_data,partyabbre %in%  top_party_name$partyabbre)
top_parties$partyabbre <- factor(top_parties$partyabbre)

#Plotting top parties vs no of candidates 
ggplot(top_parties, aes(x = reorder(partyabbre, partyabbre, function(x) length(x)))) + 
  geom_histogram(stat = 'count',fill = '#C51162') +
  geom_text(stat = "count",
            color = "white",
            vjust = 0.35, 
            hjust = 1.2, 
            size = 5,
            aes(y = ..count.., label = ..count..)) +
  theme(axis.text.x = element_text(angle = 90, size = 9)) +
  coord_flip() + 
  theme_minimal() +
  xlab("Political Parties") + 
  ylab("candidates count") +
  ggtitle("Top Parties Winning the elections")


```


We can see that out of top ten parties two parties have the largest share which are the **BJP** and **INC**.

```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots_4}
#getting summary of parties by years
by(top_parties$partyabbre,top_parties$year,summary)

#plooting stacked bar graph with top parties by year
ggplot(data = top_parties, aes(x = year, fill = partyabbre)) +
  geom_bar(stat = 'count',color = "black") + 
  coord_flip() + 
  xlab('Years') + 
  ylab("Parliament seats won by political parties") + 
  ggtitle('Rise and Fall of top parties')
```

